% Springer Nature (sn-jnl) version, for submission to Mathematical Geosciences.
%
% REQUIRES the official Springer Nature LaTeX template, NOT on CTAN/TeX Live:
%   download "LaTeX template (zip)" from
%   https://www.springernature.com/gp/authors/campaigns/latex-author-support
%   (or the Overleaf "Springer Nature article" template), and place in this
%   folder at least:  sn-jnl.cls  and the reference style sn-basic.bst
%   the reference style sn-mathphys-ay.bst (Math & Physical Sciences,
%   author-year --- the style used by Mathematical Geosciences).
%
% The section files in sections/ are class-agnostic and are shared verbatim
% with main.tex (the generic article draft that compiles without the class).
\documentclass[pdflatex,sn-mathphys-ay]{sn-jnl}

% sn-jnl loads amsmath, amsthm, graphicx, hyperref, natbib (but not amssymb).
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}

% theorem-like environments
\theoremstyle{definition}
\newtheorem{property}{Property}
\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

% notation shortcuts (shared with main.tex)
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\norm}[1]{\lVert #1 \rVert}
\newcommand{\neff}{n_{\mathrm{eff}}}
\newcommand{\dip}{\mathrm{dip}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\begin{document}

\title[ESS-calibrated test of cluster homogeneity]{%
Effective-sample-size calibration of a modality test for cluster homogeneity
and the number of classes in spatially autocorrelated data}

\author[1]{\fnm{Francisco} \sur{Parra}}\email{francisco.parra.o@usach.cl}
\affil[1]{\orgname{Universidad de Santiago de Chile},
  \orgaddress{\city{Santiago}, \country{Chile}}}

\abstract{%
Selecting the number of clusters $K$ is a basic step in unsupervised
classification. The standard internal criteria (the gap statistic, the
average silhouette, the Calinski--Harabasz and Davies--Bouldin indices)
assume independent observations, and so are miscalibrated on spatial data.
There, neighbouring locations are autocorrelated, the effective sample size is
far below the nominal one, and a smooth large-scale trend mimics discrete
structure. We show that the consequence is systematic over-detection:
on autocorrelated fields without discrete classes these criteria report
spurious clusters in every one of $75$ controlled realisations, and on real
imagery they fragment single land-cover regions. We recast the choice of $K$ as a
recursive homogeneity test: at each step, whether a region is a single class or
carries multimodality in excess of spatial autocorrelation. The
proposed method, ESS-Dip, evaluates the Hartigan dip statistic along the
locally optimal split direction. It calibrates the statistic against the
effective sample size derived from a within-class autocorrelation range, which
a local declustering estimator separates from the between-class scale; a
low-order trend is removed adaptively. A recursive partitioning returns $K$, and the test
carries an asymptotic false-split-control property. On a factorial simulation
with known ground truth, ESS-Dip reaches unit specificity with competitive
recovery; its recall-oriented variant ESS-Dip-R, calibrated against the model's
Gaussian null, attains the best balanced score of fifteen criteria at the same
specificity. On the Indian Pines, Salinas and Sentinel-2 scenes both identify
single land-cover regions as homogeneous where the standard criteria do not. The method is conservative by design, and its negatives are reliable:
when it reports a single class that verdict can be trusted, suiting ESS-Dip to
use as a homogeneity guard or a stopping rule rather than as an exhaustive
enumerator of weakly separated or strongly imbalanced classes, a trade-off we
characterise. Code and data are released.}

\keywords{number of clusters, cluster validity, spatial autocorrelation,
effective sample size, declustering, unsupervised classification,
remote sensing}

\maketitle

\input{sections/01_introduction}
\input{sections/02_background}
\input{sections/03_formulation}
\input{sections/04_simulation}
\input{sections/05_application}
\input{sections/06_discussion}
\input{sections/07_conclusions}
\input{sections/08_availability}

\bibliography{refs}

\end{document}
