% Manuscript — target: Mathematical Geosciences (Springer / IAMG)
% Generic article preamble for drafting. Switch to Springer's svjour3/sn-jnl
% class at submission stage ("your paper your way" allows generic format first).
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage[round]{natbib}
\usepackage{hyperref}

% theorem-like environments
\theoremstyle{definition}
\newtheorem{property}{Property}
\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}
\theoremstyle{remark}
\newtheorem{remark}{Remark}

% notation shortcuts
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\norm}[1]{\lVert #1 \rVert}
\newcommand{\neff}{n_{\mathrm{eff}}}
\newcommand{\dip}{\mathrm{dip}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\title{Effective-sample-size calibration of a modality test for cluster
homogeneity and the number of classes in spatially autocorrelated data}
\author{F. Parra \and [co-authors]}
\date{Draft, \today}

\begin{document}
\maketitle

\begin{abstract}
Selecting the number of clusters $K$ is a basic step in unsupervised
classification. The standard internal criteria (the gap statistic, the
average silhouette, the Calinski--Harabasz and Davies--Bouldin indices)
assume independent observations, and so are miscalibrated on spatial data.
There, neighbouring locations are autocorrelated, the effective sample size is
far below the nominal one, and a smooth large-scale trend mimics discrete
structure. We show that the consequence is systematic over-detection:
on autocorrelated fields without discrete classes these criteria report
spurious clusters in every one of $75$ controlled realisations, and on real
imagery they fragment single land-cover regions. We recast the choice of $K$ as a recursive homogeneity test: at each step,
whether a region is a single class or carries \emph{multimodality in excess of
spatial autocorrelation}. The proposed method, ESS-Dip, evaluates the Hartigan dip
statistic along the locally optimal split direction. It calibrates the statistic
against the effective sample size derived from a within-class autocorrelation
range, which a local declustering estimator separates from the between-class
scale; a low-order trend is removed adaptively. A recursive partitioning returns $K$,
and the test carries an asymptotic false-split-control property. On a factorial
simulation with known ground truth, ESS-Dip reaches unit specificity with
competitive recovery; its recall-oriented variant ESS-Dip-R, calibrated against
the model's Gaussian null, attains the best balanced score of fifteen criteria
at the same specificity. On the Indian Pines, Salinas and Sentinel-2 scenes both
identify single land-cover regions as homogeneous where the standard criteria
do not. The method is conservative by design, and its negatives are reliable:
when it reports a single class that verdict can be trusted, suiting ESS-Dip to
use as a homogeneity guard or a stopping rule rather than as an exhaustive
enumerator of weakly separated or strongly imbalanced classes, a trade-off we
characterise. Code and data are released.
\end{abstract}

\noindent\textbf{Keywords:} number of clusters \textperiodcentered\ cluster
validity \textperiodcentered\ spatial autocorrelation \textperiodcentered\
effective sample size \textperiodcentered\ declustering \textperiodcentered\
unsupervised classification \textperiodcentered\ remote sensing

\input{sections/01_introduction}
\input{sections/02_background}
\input{sections/03_formulation}
\input{sections/04_simulation}
\input{sections/05_application}
\input{sections/06_discussion}
\input{sections/07_conclusions}
\input{sections/08_availability}

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
