\section{Introduction}\label{sec:intro}

Unsupervised classification partitions a data set into groups without labelled
examples, and in doing so it must answer a prior question: how many groups are
there? The choice of the number of clusters $K$ governs every downstream
interpretation, and in the analysis of remotely sensed imagery, where $K$ is
the number of land-cover classes an analyst will map, it is made routinely
and consequentially. A mature toolkit exists to make it: the gap statistic
\citep{Tibshirani2001}, the average silhouette \citep{Rousseeuw1987}, the
Calinski--Harabasz \citep{CalinskiHarabasz1974} and Davies--Bouldin
\citep{DaviesBouldin1979} indices each score a partition by its internal
geometry and select the $K$ that optimises the score.

These criteria share a hidden assumption: that the observations are
independent. Their reference levels, the uniform null of the gap statistic and
the implicit baseline of a silhouette, describe what the geometry would look
like were the samples exchangeable draws. Spatial data are not exchangeable.
Neighbouring pixels are autocorrelated \citep{Tobler1970}, so the effective
number of independent observations is far below the nominal pixel count
\citep{Cressie1993}, and a smooth illumination or topographic gradient spreads
the feature cloud along a manifold that an independence null cannot
distinguish from genuine groups. The consequence, which we quantify in
Sections~\ref{sec:simulation}--\ref{sec:application}, is systematic
\emph{over-detection}. On synthetic fields containing no discrete classes the
gap statistic reports on average between seven and eight; on real imagery
the silhouette, Calinski--Harabasz and Davies--Bouldin indices fragment a
single, homogeneous crop field into several sub-classes in every instance we
examined. A criterion that cannot recognise the absence of clusters is a
precarious foundation for deciding their number.

Correcting this is not a matter of swapping in a better null. Because the
within-cluster dispersion that these criteria minimise is a second-order
quantity, a reference matched to the data's second-order spatial structure,
its variogram, reproduces the same dispersion and explains the clusters
away. Distinguishing real classes from autocorrelation therefore requires
information beyond second order, namely the \emph{multimodality} of the feature
distribution; and distinguishing real classes from a smooth trend requires
separating a continuous drift from discrete structure, a separation the
geostatistical decomposition into trend and residual makes natural. The idea of
calibrating a clustering criterion against an autocorrelation-aware null is
itself not new. \citet{HennigLin2015} do exactly this for discrete
presence--absence data, but it has not been carried to the continuous-field
setting of gridded imagery, nor has the confounding of trend with class been
addressed in selecting $K$.

This paper supplies both. We recast the question \emph{how many classes?} as a
sequence of homogeneity decisions: at each node we test whether a region is a
single class or instead carries \emph{multimodality in excess of spatial
autocorrelation}, and make that test valid on autocorrelated fields by
calibrating it at the effective sample size. The
construction rests on three geostatistical devices: the effective sample size
itself \citep{Cressie1993,Dutilleul1993}, evaluated at the indicator integral
range appropriate to a functional of the distribution rather than the field
integral range that governs the mean (Section~\ref{subsec:sgs}), which sets the
calibration; a
\emph{local} estimate of the within-class autocorrelation range, in the manner
of declustering \citep{IsaaksSrivastava1989,Deutsch1998}, which prevents the
between-class mosaic from corrupting that calibration; and an adaptive trend--residual
decomposition, which removes a smooth drift only when it is statistically
resolvable. A recursive partitioning driven by the calibrated dip statistic
\citep{HartiganHartigan1985} returns $K$. Because the test is tuned for
specificity, its negatives are trustworthy: when it declares a region
homogeneous that verdict can be relied on, which makes ESS-Dip as useful as a
homogeneity guard or a stopping rule for region-growing as it is as a
conservative estimator of $K$. Our contributions are:

\begin{itemize}
  \item We document, on controlled ground truth, that the standard
    number-of-clusters criteria over-detect under spatial autocorrelation
    without exception, failing to recognise class-free scenes in every one of
    $75$ realisations (Section~\ref{sec:simulation}).
  \item We propose \emph{ESS-Dip}, an effective-sample-size-calibrated test of
    cluster homogeneity (whether a region holds discrete structure beyond its
    spatial autocorrelation), combining a within-class range from local
    range estimation with adaptive trend removal (Section~\ref{sec:method}); applied
    recursively it returns a conservative $K$, and we state an asymptotic
    false-split-control property (Property~\ref{prop:fpr}) that the method
    confirms empirically; a recall-oriented variant, ESS-Dip-R, trades the
    distribution-free uniform null for the model's Gaussian null to recover more
    structure at the same specificity.
  \item We evaluate fifteen criteria on a factorial simulation with known $K$,
    including the calibrated-validity-index method of \citet{HennigLin2015} that
    ESS-Dip generalises; ESS-Dip reaches unit specificity, its recall variant
    ESS-Dip-R attains the best balanced score at the same specificity, and an
    ablation attributes the gain to the local range and the adaptive detrending
    (Section~\ref{sec:simulation}).
  \item We confirm the behaviour on three real scenes spanning hyperspectral
    and multispectral sensing (Indian Pines, Salinas, and a Sentinel-2 scene
    labelled by ESA WorldCover), where ESS-Dip identifies single land-cover
    regions as homogeneous where the standard criteria do not
    (Section~\ref{sec:application}).
\end{itemize}

The remainder of the paper is organised as follows.
Section~\ref{sec:background} reviews number-of-clusters criteria, the
geostatistical notions of effective sample size and declustering, and
modality-based clustering, and positions the contribution.
Section~\ref{sec:method} develops the method and its calibration.
Sections~\ref{sec:simulation} and~\ref{sec:application} report the simulation
study and the application to land-cover imagery. Section~\ref{sec:discussion}
discusses the precision--recall trade-off and limitations, and
Section~\ref{sec:conclusions} concludes. Code and data are released to
reproduce every result.
