\section{Mathematical formulation}\label{sec:method}

We formulate the selection of the number of clusters as a sequence of
hypothesis tests for \emph{multimodality in excess of spatial
autocorrelation}. Section~\ref{subsec:setup} fixes notation;
Section~\ref{subsec:null} states the null against which a candidate split is
judged; Sections~\ref{subsec:dip}--\ref{subsec:detrend} develop the test
statistic, its effective-sample-size calibration, the within-class range
estimator, and the trend--residual decomposition; Section~\ref{subsec:algo}
assembles these into a recursive partitioning algorithm;
Section~\ref{subsec:prop} states the resulting false-split control; and
Section~\ref{subsec:complexity} discusses computation.

\subsection{Setting and notation}\label{subsec:setup}

Let $D\subset\Z^2$ be a finite regular lattice (the image domain) with
$N=\lvert D\rvert$ pixels, and let
\[
  \bm z:D\to\R^p,\qquad
  \bm z(s)=\bigl(z_1(s),\dots,z_p(s)\bigr)^{\!\top},
\]
be a multivariate field assigning to each site $s\in D$ a $p$-dimensional
feature (e.g.\ a spectral signature). Writing $\bm z$ as $N$ feature vectors
$\{\bm z(s)\}_{s\in D}$, unsupervised classification seeks a partition of $D$
into $K$ groups whose feature vectors are internally homogeneous. Our object
of interest is the \emph{number} of groups $K$, not the partition itself.

We treat $\bm z$ as one realisation of a second-order stationary random field
and decompose it, in the geostatistical manner, into a deterministic trend and
a stochastic residual,
\begin{equation}\label{eq:decomp}
  \bm z(s)=\bm m(s)+\bm\varepsilon(s),
\end{equation}
where $\bm m$ is a smooth large-scale drift (illumination, topography,
atmospheric gradients) and $\bm\varepsilon$ is a zero-mean residual carrying
the short-range autocorrelation and any discrete (between-class) structure.
The second-order dependence of $\bm\varepsilon$ is summarised by its
correlogram $\rho(\bm h)$, or equivalently its variogram, and condensed into a
single scalar by the \emph{decorrelation area}, the integral range of
\citet{Cressie1993},
\begin{equation}\label{eq:area}
  a=\iint \rho(\bm h)\,d\bm h,
\end{equation}
the area occupied by one spatially independent observation. Estimating $a$ from
a fitted variogram is the subject of Section~\ref{subsec:range}; for a subset
$C\subseteq D$ we write $n_C=\lvert C\rvert$.

\subsection{Clusters as excess multimodality}\label{subsec:null}

Internal validity criteria for $K$, namely the gap statistic
\citep{Tibshirani2001}, the average silhouette, Calinski--Harabasz and
Davies--Bouldin, compare a within-cluster dispersion against the value
expected with \emph{no} clustering, under an \emph{independent} reference
distribution. Two features of spatial data invalidate this comparison. First,
autocorrelation reduces the number of independent observations far below $N$,
so dispersion-based statistics are evaluated as if the sample were much larger
than its information content \citep{Cressie1993,Dutilleul1993}. Second, a
smooth drift $\bm m$ spreads the feature cloud along a low-dimensional manifold
and is read as structure by an independence null. The combined effect is a
systematic \emph{over-detection} of clusters, which we document empirically in
Section~\ref{sec:simulation}.

We therefore replace the dispersion comparison by a test of \emph{modality}.
Discrete classes manifest as a \emph{multimodal} feature distribution; a
single class (however autocorrelated, and however smoothly it varies in
space) is \emph{unimodal}. This motivates the reference hypothesis under
which a candidate cluster is judged homogeneous.

\begin{assumption}[Unimodal Gaussian-random-field null]\label{ass:null}
Under the null $H_0$ a cluster's residual field $\bm\varepsilon$ restricted to
$C$ is a second-order stationary Gaussian random field with decorrelation area
$a$ and \emph{no} discrete component. In particular the marginal law of any
fixed linear projection $\bm u^{\!\top}\bm\varepsilon$, $\bm u\in\R^p$, is
univariate Gaussian, hence unimodal.
\end{assumption}

Assumption~\ref{ass:null} is the continuous-field counterpart of the
autocorrelation-aware null of \citet{HennigLin2015}, who calibrated a
validity index against a non-clustering reference for point data; here the
reference is a Gaussian random field, the maximum-entropy unimodal model given
the second-order structure, of the kind routinely generated by sequential
Gaussian simulation \citep{Goovaerts2004,WagnerDray2015}.

\subsection{The dip statistic along the split direction}\label{subsec:dip}

Modality is quantified by Hartigan's dip \citep{HartiganHartigan1985}. For a
univariate sample with empirical distribution function $F_m$ (size $m$), the
dip is the supremum distance to the closest unimodal distribution,
\begin{equation}\label{eq:dip}
  \dip(F_m)=\inf_{G\in\mathcal U}\ \sup_{x\in\R}\,\lvert F_m(x)-G(x)\rvert,
\end{equation}
$\mathcal U$ being the class of unimodal distribution functions. The dip is
$0$ for a perfectly unimodal sample and grows with the separation between
modes.

Testing \eqref{eq:dip} on a one-dimensional projection requires a direction
along which a genuine bimodality, if present, is exposed. Rather than the
leading principal axis, which can hide modes spread across several
dimensions, we test along the direction that a tentative binary split
proposes. Given a cluster $C$, let $k$-means with $k=2$ return centroids
$\bm\mu_0,\bm\mu_1$, and set the unit split direction and projected sample
\begin{equation}\label{eq:proj}
  \bm u=\frac{\bm\mu_1-\bm\mu_0}{\norm{\bm\mu_1-\bm\mu_0}},
  \qquad
  y(s)=\bm u^{\!\top}\bm z(s),\quad s\in C .
\end{equation}
The statistic is $\hat D_C=\dip(\hat F_C)$, where $\hat F_C$ is the empirical
distribution function of $\{y(s)\}_{s\in C}$. Coupling the test to the split
it must license makes the procedure sensitive to the bimodality $k$-means
detects, while preserving its calibration under $H_0$: projecting a unimodal
(Gaussian) cloud onto any fixed direction returns a unimodal sample,
regardless of where $k$-means places the cut.

\subsection{Effective-sample-size calibration}\label{subsec:ess}

The null distribution of the dip depends on the sample size: for $m$
\emph{independent} draws from the least-favourable unimodal law (the uniform;
\citealp{HartiganHartigan1985}), $\dip(F_m)=O_p(m^{-1/2})$, and we denote by
$q_{1-\alpha}(m)$ its $(1-\alpha)$ quantile. With autocorrelated data the
nominal size $n_C$ overstates the information content: the empirical
distribution of $n_C$ correlated values fluctuates about the population
distribution at the rate of
\begin{equation}\label{eq:neff}
  \neff=\frac{n_C}{a}
\end{equation}
\emph{independent} values, the effective sample size of a field with
decorrelation area $a$ \citep{Cressie1993,Dutilleul1993}. We therefore
calibrate the observed dip against the null at the effective, not the nominal,
size: the test rejects unimodality of $C$, and thereby licenses a split,
when
\begin{equation}\label{eq:test}
  \hat D_C \;>\; q_{1-\alpha}\!\bigl(\,\lfloor\neff\rceil\,\bigr),
\end{equation}
with $\lfloor\cdot\rceil$ the nearest integer. The quantile
$q_{1-\alpha}(\cdot)$ has no closed form and is obtained by Monte Carlo
($\dip$ of uniform samples of size $\lfloor\neff\rceil$, cached per size bucket),
so the calibration adds negligible cost. Autocorrelation thus raises the
significance threshold (through a smaller $\neff$) without discarding data: the
statistic $\hat D_C$ is still computed from all $n_C$ pixels, retaining full
detection power.

\subsection{Within-class decorrelation area by local variogram fitting}
\label{subsec:range}

The calibration \eqref{eq:neff} requires the decorrelation area \eqref{eq:area}
of the \emph{within-class} residual $\bm\varepsilon$, not of the between-class
mosaic. Estimating it from the whole scene conflates the two: large contiguous
classes inflate the apparent range, $a$ grows, $\neff$ collapses, and the test
loses all power, the field-scale rather than the within-class autocorrelation
being measured. We avoid this conflation by estimating $a$ \emph{locally}, in
the spirit of declustering corrections for spatially clustered support
\citep{IsaaksSrivastava1989,Deutsch1998}.

Partition $D$ into square tiles $\{T_j\}$ of side $t$; most tiles fall
\emph{within} a single class, so their second-order structure reflects
$\bm\varepsilon$ alone. For each tile we form the radially averaged empirical
semivariogram $\hat\gamma_j(h)$ of the first principal component of
$\{\bm z(s)\}_{s\in T_j}$ (obtained in $O(t^2\log t)$ from the fast Fourier
transform autocovariance) and fit, by weighted least squares, the
best of an exponential, spherical and Gaussian model with nugget $c_{0}$,
partial sill $c$ and range parameter $a_j$. For an isotropic model the integral
range \eqref{eq:area} is available in closed form,
\begin{equation}\label{eq:intrange}
  a_j \;=\; 2\pi\,\frac{c}{c_{0}+c}\!\int_0^\infty\! r\,\rho(r)\,dr
       \;=\; \kappa\,\frac{c}{c_{0}+c}\,a_j^{2},
\end{equation}
with $\kappa=2\pi,\ \pi,\ 0.2\pi$ for the exponential, Gaussian and spherical
models respectively; the nugget fraction $c_{0}/(c_{0}+c)$, being uncorrelated,
enlarges the effective sample size. The within-class decorrelation area is the
tile median,
\begin{equation}\label{eq:areamed}
  a=\operatorname{median}_j\,a_j ,
\end{equation}
robust to the minority of tiles that straddle a class boundary. The constant
$\kappa$ is exactly what an isotropic $1/e$-range proxy $a\approx\ell^2$ leaves
implicit; the variogram makes it explicit and model-dependent, which matters
because $\kappa$ rescales $\neff$ and hence the calibration. We return to the
sensitivity of the test to this constant, and to an assumption-free
alternative, in Section~\ref{subsec:sgs}.

\subsection{Adaptive trend--residual separation}\label{subsec:detrend}

A smooth drift $\bm m$ is itself unimodal yet, left in place, enlarges $\ell$
and biases the test. We remove it adaptively. For each band $b$ fit a
low-order polynomial surface $\hat m_b(s)=\bm\beta_b^{\!\top}\bm\phi(s)$ by
ordinary least squares, where $\bm\phi$ collects the monomials in the pixel
coordinates up to degree two, and let
\begin{equation}\label{eq:r2}
  R^2_{\mathrm{poly}}
  =\frac1p\sum_{b=1}^{p}
   \left(1-\frac{\sum_{s}\bigl(z_b(s)-\hat m_b(s)\bigr)^2}
               {\sum_{s}\bigl(z_b(s)-\bar z_b\bigr)^2}\right)
\end{equation}
be the average variance explained by the drift. A genuine large-scale trend
produces $R^2_{\mathrm{poly}}$ close to one, whereas a discrete mosaic, not
well represented by a low-order polynomial, produces a small value. We
therefore test the residual $\bm\varepsilon=\bm z-\hat{\bm m}$ in place of
$\bm z$ whenever
\begin{equation}\label{eq:detrendrule}
  R^2_{\mathrm{poly}}>\tau,
\end{equation}
and test $\bm z$ unchanged otherwise. Removing the drift only when it is
present protects the power of the test on trend-free scenes, where detrending
would needlessly strip signal. Equation~\eqref{eq:decomp} together with
\eqref{eq:detrendrule} is the universal-kriging decomposition restricted to
the regime in which the drift is statistically resolvable.

\subsection{A Gaussian-simulation reference}\label{subsec:sgs}

The effective-sample-size calibration \eqref{eq:test} is an approximation: it
maps an autocorrelated sample of size $n_C$ to an independent sample of size
$\neff$ and reads the dip quantile there. Two caveats attend it. First, the
integral range \eqref{eq:intrange} is derived for the variance of the sample
\emph{mean}; the dip is a functional of the empirical \emph{distribution}, for
which the appropriate effective size carries a different constant, so $\kappa$
is best regarded as a calibration parameter rather than a fixed geometric
factor. Second, it presumes isotropy. A reference that avoids both is the
geostatistical \emph{neutral model} \citep{Goovaerts2004}: simulate, by the
spectral (fast-Fourier) method \citep{DietrichNewsam1993,Lantuejoul2002}, an
ensemble of unconditional Gaussian realisations
$\{\tilde y^{(b)}\}_{b=1}^{B}$ on the support of $C$ whose covariance is the
fitted variogram of the projection $y$, and reject unimodality when
\begin{equation}\label{eq:sgstest}
  \hat D_C \;>\; \mathrm{Quantile}_{1-\alpha}\bigl\{\dip(\tilde y^{(b)})\bigr\}_{b=1}^{B}.
\end{equation}
For a \emph{single} homogeneity decision on a full rectangular domain,
testing $K=1$ against $K\ge2$, this Monte Carlo test makes no independence
approximation and accommodates anisotropic or nested structure, and we show in
Section~\ref{subsec:geostat-ref} that it decides correctly where the
effective-sample-size shortcut requires its calibration constant. Two
limitations keep it from supplanting \eqref{eq:test} as the estimator, however.
It is exact only up to the \emph{plug-in} variogram, fitted from the same data;
and when iterated over the recursion its per-node variogram must be fitted on
the non-rectangular support of a candidate cluster, and the repeated tests
incur multiple comparisons. We implement the resulting recursive estimator
(Section~\ref{subsec:geostat-ref}): a direct version over-splits, a
multiple-testing correction makes it competitive with \eqref{eq:test} on
controlled fields, but its plug-in variogram on irregular supports keeps it
behind on real imagery, and at the cost of $B$ simulations per node against the
cached, near-linear cost of \eqref{eq:test}. We therefore retain
\eqref{eq:test} as the method and use \eqref{eq:sgstest} as an
assumption-light check on the leading split.

\subsection{Recursive partitioning}\label{subsec:algo}

The components above are combined into a top-down recursion
(Algorithm~\ref{alg:essdip}). Beginning with $C=D$, a cluster is split into
its $2$-means children whenever the calibrated dip test \eqref{eq:test} is
significant and both children retain at least one decorrelation patch
($n\ge a$, so that $\neff\ge1$); otherwise it becomes a leaf. The estimate
$\hat K$ is the number of leaves. To absorb the variability of the $k$-means
initialisation the recursion is repeated $B$ times with independent seeds and
the median of $\{\hat K^{(b)}\}_{b=1}^B$ is reported.

\begin{algorithm}[t]
\caption{ESS-calibrated modality test for $K$ (single ensemble member)}
\label{alg:essdip}
\begin{algorithmic}[1]
\Require field $\bm z$ on $D$; level $\alpha$; degree-2 drift threshold $\tau$;
         tile side $t$; minimum effective size $\nu$
\If{$R^2_{\mathrm{poly}}(\bm z)>\tau$}\Comment{Eq.~\eqref{eq:r2}--\eqref{eq:detrendrule}}
  \State $\bm z\gets \bm z-\hat{\bm m}$\Comment{trend removal}
\EndIf
\State $a\gets\operatorname{median}_j a_j$ \Comment{local integral range, Eq.~\eqref{eq:intrange}--\eqref{eq:areamed}}
\State stack $\gets\{D\}$;\quad leaves $\gets 0$
\While{stack $\neq\emptyset$}
  \State pop $C$ from stack;\quad $\neff\gets n_C/a$
  \If{$\neff<\nu$}\State leaves $\gets$ leaves $+1$;\quad \textbf{continue}
       \Comment{insufficient independent information}
  \EndIf
  \State $(\bm\mu_0,\bm\mu_1)\gets$ \textsc{$2$-means}$(C)$;\quad
         $\bm u\gets(\bm\mu_1-\bm\mu_0)/\norm{\bm\mu_1-\bm\mu_0}$
  \State $\hat D_C\gets\dip\bigl(\{\bm u^{\!\top}\bm z(s)\}_{s\in C}\bigr)$
         \Comment{Eq.~\eqref{eq:dip}--\eqref{eq:proj}}
  \If{$\hat D_C>q_{1-\alpha}(\lfloor\neff\rceil)$ \textbf{and} both children
      have $n\ge a$}
     \State push both $2$-means children onto stack\Comment{Eq.~\eqref{eq:test}}
  \Else
     \State leaves $\gets$ leaves $+1$
  \EndIf
\EndWhile
\State \Return $\hat K=$ leaves
\end{algorithmic}
\end{algorithm}

\subsection{False-split control}\label{subsec:prop}

The calibration \eqref{eq:test} is designed so that, in the absence of
discrete structure, the recursion does not split. We state this as a heuristic
property of the per-node test, made exact by the simulation reference
\eqref{eq:sgstest}, whose consequence for $\hat K$ follows by induction over
the (finite) recursion.

\begin{property}[Heuristic false-split control]\label{prop:fpr}
Let a node $C$ satisfy Assumption~\ref{ass:null}, and let $\neff=n_C/a\to\infty$
with $a$ fixed. Then, to the order of the block-independence approximation
below, the probability that the test \eqref{eq:test} rejects unimodality is at
most $\alpha+o(1)$; consequently, if every node of the recursion satisfies
Assumption~\ref{ass:null} (the scene contains a single class), then
$\Prob(\hat K=1)\to1$.
\end{property}

\noindent\textit{Heuristic argument.}\;
Under Assumption~\ref{ass:null} the projected field $y=\bm u^{\!\top}
\bm\varepsilon$ has, for any fixed $\bm u$, a Gaussian marginal, so its
population dip is zero. We approximate the information content of the $n_C$
spatially dependent values in $\hat F_C$ by that of $\neff$ independent ones:
partitioning $C$ into $\approx\neff$ blocks of area $a$, one effectively
independent observation per block, the fluctuations of $\hat F_C$ about the
Gaussian population law have, to first order, the magnitude of an
$\neff$-sample empirical process \citep{Cressie1993,Dutilleul1993}. Then
$\hat D_C$ is approximately distributed as $\dip(F_{\neff})$ under independence,
and comparison with the matched quantile $q_{1-\alpha}(\lfloor\neff\rceil)$
yields rejection probability $\alpha+o(1)$. The split direction $\bm u$ is
data-dependent, but under $H_0$ the cloud is unimodal in every direction, so
optimising the cut over $\bm u$ does not create a mode.

We label this a property rather than a theorem deliberately. A rigorous
statement would require an empirical-process central limit theorem for the dip
functional under a spatial mixing condition, together with control of the
block-independence constant in \eqref{eq:intrange}, which we do not pursue. The
Gaussian-simulation reference \eqref{eq:sgstest} controls the level exactly for
a single homogeneity test and so provides an assumption-light check on the
leading split, but, as Section~\ref{subsec:geostat-ref} shows, it does not yield
a better recursive estimator. The effective-sample-size test \eqref{eq:test} is
thus an approximation whose adequacy is assessed empirically:
Section~\ref{sec:simulation} finds that, over the $75$ null and trend
realisations with no discrete classes, the procedure returns $\hat K=1$ in
every case, whereas the gap statistic and the silhouette, Calinski--Harabasz
and Davies--Bouldin indices report spurious clusters in all of them.

\subsection{Computational considerations}\label{subsec:complexity}

The range estimator \eqref{eq:areamed} costs one fast Fourier transform per
tile, $O(N\log N)$ in total. Each visited node runs one $2$-means fit, of cost
$O(n_C p)$ with mini-batch updates, and one dip evaluation, $O(n_C\log n_C)$
after sorting the projection; the Monte Carlo quantiles
$q_{1-\alpha}(\cdot)$ are computed once per effective-size bucket and reused.
As the recursion visits each pixel in $O(\hat K)$ nodes, one ensemble member is
near-linear in $N$, and the $B$ members are independent and trivially parallel.
The procedure therefore scales to full scenes;
Section~\ref{subsec:timing} reports timings confirming a near-linear empirical
exponent and a memory footprint a small multiple of the data, together with a
constant-factor advantage over the gap statistic, whose reference replicates
multiply its cost.
