\section{Discussion}\label{sec:discussion}

\paragraph{A precision-oriented criterion.}
The results of Sections~\ref{sec:simulation} and~\ref{sec:application} draw a
consistent picture across synthetic fields, hyperspectral scenes and
multispectral imagery: standard internal criteria are powerful at recovering
well-separated classes but cannot recognise their absence, whereas ESS-Dip
controls the false-split rate at the cost of recall when classes are many or
weakly separated. The two behaviours are not interchangeable failures of
calibration but the two ends of a deliberate trade-off. By referring the dip
statistic to the effective rather than the nominal sample size, the test is
made conservative in exact proportion to the information the autocorrelation
removes; the level $\alpha$ and the range $\ell$ are the knobs that set where on
the precision--recall curve the method sits. We have fixed them once, without
per-scene tuning, and the resulting operating point is one of high precision:
ESS-Dip answers reliably whether discrete structure is present \emph{beyond}
spatial autocorrelation, and is correspondingly cautious about enumerating
structure that the data do not independently support. For the analyst this is
the useful direction of error, a guard against the over-segmentation that
Section~\ref{subsec:realspec} documents for the standard indices, but it is
a genuine limitation when an exhaustive class inventory is the goal. This fixes
the method's natural role: not a stand-alone classifier (on real multi-class
scenes it under-reports), but a calibrated homogeneity test, a guard placed
before an unsupervised classification or a stopping rule for a region-growing
or regionalization procedure (Section~\ref{subsec:bg-constrained}), where a
negative one can trust is worth more than an exhaustive but unreliable class
count.

\paragraph{Relation to geostatistical null models.}
Assumption~\ref{ass:null} replaces the independence reference of the classical
criteria by a unimodal Gaussian random field, the continuous-field counterpart
of the spatial neutral models used in disease mapping \citep{Goovaerts2004} and
of Moran spectral randomization \citep{WagnerDray2015}. We realise it
implicitly, through the effective-sample-size calibration of the dip, rather
than by simulating reference fields. A more explicit route would calibrate the
statistic against an ensemble of sequential-Gaussian-simulation realisations
conditioned on the empirical variogram; this would relax the isotropy built
into our scalar range $\ell$ and accommodate anisotropic or nested
autocorrelation structures at the cost of the simulation effort. We see the
present scalar calibration and a full simulation-based null as two points on a
spectrum trading fidelity against computation, and the near-linear cost of the
former (Section~\ref{subsec:complexity}) as its principal practical advantage
on megapixel scenes.

\paragraph{Limitations.}
Three limitations bound the method's applicability. First, when many spectrally
overlapping classes coexist, as in the sixteen-class hyperspectral scenes, no
unsupervised criterion, ours included, approaches the nominal class count
(Section~\ref{subsec:realfull}); recovering such inventories requires
supervision or auxiliary information that an internal criterion cannot supply.
Second, strong class imbalance degrades recall: when one class dominates the
scene, as the crop class does in the Sentinel-2 image, the leading split
direction is governed by that majority and minority classes are not resolved.
Third, the false-split control of Property~\ref{prop:fpr} is an asymptotic
statement under a block-independence idealisation, not a finite-sample
guarantee, and the projection onto the $2$-means direction, though effective in
our experiments, is a heuristic rather than an optimal choice.

\paragraph{Extensions.}
Each limitation suggests a direction. A recall-oriented variant could revisit
leaves at a relaxed level, or test minority modes against the residual of the
majority, to mitigate imbalance without sacrificing the specificity of the
first pass. Replacing the scalar range by an anisotropic variogram model, and
the uniform dip null by conditional simulation, would extend the calibration to
non-stationary autocorrelation, although Section~\ref{subsec:aniso} shows the
isotropic estimate is already robust to directional autocorrelation up to an
eight-fold anisotropy ratio. The same effective-sample-size
correction applies, unchanged, to spatio-temporal stacks, where the
decorrelation ``area'' becomes a space--time volume. Finally, because our
contribution is a stopping rule rather than a partitioning constraint, it
composes naturally with spatially constrained clustering
(Section~\ref{subsec:bg-constrained}): the calibrated test could decide when a
regionalization has produced as many regions as the data independently
support.
